A comparison of model-based and design-based approaches for spatial data.

### 2021-06-07 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
minor typo fixes in initial draft.

Several comparisons between model-based and design-based for spatial data have been considered, but they tend to compare spatial models to IRS or SIRS (stratified IRS). No comparison has been made between spatial models and spatially balanced design-based approaches.

@@

On the other hand, model-based inference imposes additional assumptions on the data with a potential to provide more precise estimates if the additional assumptions hold. Instead of estimating true but unknown parameters, the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. This is a fundamental philosophical difference between sampling-based and model-based approaches. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all sites, we would have an exact prediction for the mean of our one realized spatial surface, without any uncertainty. But, the true mean of the spatial process that generated our realized data is still not known, and, in the prediction context, we typically do not care much about what value the mean of the underlying process takes.

@@

Spatially balanced sampling algorithms use spatial information to obtain samples spread out in space. Spatially balanced samples are useful because they tend to yield estimators that are more precise than estimators constructed from an sampling algorithm that is not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]. To quantify spatial balance, @stevens2004spatially proposed statistics based on Voroni polygons. Many spatially balanced sampling algorithms exist, including the Generalized Random Tessellation Stratified [@stevens2004spatially], the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. Here we focus on the Generalized Random Tessellation Stratified (GRTS) algorithm, which has several attractive properties that we discuss next.

The GRTS algorithm is used to sample from finite and infinite sample frames. A finite sample frames contains a finite number of sampling units and is related to a point geometry. An infinite sample frame contains an infinite number of sampling units and is related to linear and polygon geometries. Examples of point, linear, and polygon resources include lake centroids, stream networks, and wetland areas, respectively. In addition to its applicability for finite and infinite sample frames, the GRTS algorithm naturally accommodates stratified designs and designs with unequal selection probabilities. The algorithm has also been used to select replacement sites using reverse hierarchical ordering [@stevens2004spatially]. Replacement sites are used to replace sites in the original sample that cannot be sampled, often as a result of physical difficulty in reaching the site or landowner denial of access to the sites. More recently, the GRTS algorithm also accommodates legacy (historical) sites, minimum distance between sites, and nearest neighbor replacement sites. The GRTS algorithm is implemented in the \textbf{\textsf{R}} package \texttt{spsurvey} [@dumelle2021spsurvey].

@@

The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population totals (and means). If $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$, denoted by $\hat{\tau}_{HT}$, is given by
\begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Y_i \pi_i^{-1},
\end{align}
where $Y_i$ is the observed value of the $i$th sampling unit, and $\pi_i$ is the inclusion probability of the $i$th sampling unit. Estimating the variance of $\hat{\tau}_{HT}$ directly relies on knowing the probability that $s_i$ and $s_j$ are both included in the sample (second-order inclusion probability), which can be unfeasible to compute. Furthermore, this variance estimator does incorporate spatial locations. To address these challenges, [@stevens2003variance] proposed an alternative variance estimator: the local neighborhood variance estimator. The local neighborhood variance estimator does not require second-order inclusion probabilities. It also incorporates spatial locations, which tends to reduce the estimated variance of compared $\hat{\tau}_{HT}$ to a variance estimator ignoring spatial location [@stevens2003variance].

>>>>

Several comparisons between model-based and design-based for spatial data have been considered, but they tend to compare spatial models to IRS (Independent Random Sampling) or SIRS (stratified IRS). No comparison has been made between spatial models and spatially balanced design-based approaches.

@@

On the other hand, model-based inference imposes additional assumptions on the data with a potential to provide more precise estimators if the additional assumptions hold. Instead of estimating true but unknown parameters, the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. This is a fundamental philosophical difference between sampling-based and model-based approaches. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all sites, we would have an exact prediction for the mean of our one realized spatial surface, without any uncertainty. But, the true mean of the spatial process that generated our realized data is still not known, and, in the prediction context, we typically do not care much about what value the mean of the underlying process takes.

@@

Spatially balanced sampling algorithms use spatial information to obtain samples spread out in space. Spatially balanced samples are useful because they tend to yield estimators that are more precise than estimators constructed from a sampling algorithm that is not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]. To quantify spatial balance, @stevens2004spatially proposed statistics based on Voroni polygons. Many spatially balanced sampling algorithms exist, including the Generalized Random Tessellation Stratified [@stevens2004spatially], the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. Here we focus on the Generalized Random Tessellation Stratified (GRTS) algorithm, which has several attractive properties that we discuss next.

The GRTS algorithm is used to sample from finite and infinite sample frames. A finite sample frame contains a finite number of sampling units and is related to a point geometry. An infinite sample frame contains an infinite number of sampling units and is related to linear and polygon geometries. Examples of point, linear, and polygon resources include lake centroids, stream networks, and wetland areas, respectively. In addition to its applicability for finite and infinite sample frames, the GRTS algorithm naturally accommodates stratified designs and designs with unequal selection probabilities. The algorithm has also been used to select replacement sites using reverse hierarchical ordering [@stevens2004spatially]. Replacement sites are used to replace sites in the original sample that cannot be sampled, often as a result of physical difficulty in reaching the site or landowner denial of access to the sites. More recently, the GRTS algorithm also accommodates legacy (historical) sites, minimum distance between sites, and nearest neighbor replacement sites. The GRTS algorithm is implemented in the \textbf{\textsf{R}} package \texttt{spsurvey} [@dumelle2021spsurvey].

@@

The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population totals (and means). If $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$, denoted by $\hat{\tau}_{HT}$, is given by
\begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Y_i \pi_i^{-1},
\end{align}
where $Y_i$ is the observed value of the $i$th sampling unit, and $\pi_i$ is the inclusion probability of the $i$th sampling unit. Estimating the variance of $\hat{\tau}_{HT}$ directly relies on knowing the probability that $s_i$ and $s_j$ are both included in the sample (second-order inclusion probability), which can be unfeasible to compute. Furthermore, this variance estimator does not incorporate spatial locations. To address these challenges, [@stevens2003variance] proposed an alternative variance estimator: the local neighborhood variance estimator. The local neighborhood variance estimator does not require second-order inclusion probabilities. It also incorporates spatial locations, which tends to reduce the estimated variance of compared $\hat{\tau}_{HT}$ to a variance estimator ignoring spatial location [@stevens2003variance].
